plyranges: a grammar for transforming genomics data

### 2018-04-24 paper.Rmd
comment on example 2 ...
complete draft (minus final example)

---
title: "plyranges: a grammar for transforming genomics data" 
author:
  - name: Stuart Lee
    email: stuart.lee1@monash.edu
    affiliation: Monash University
  - name: Michael Lawrence
    email: michael.lawrence@gene.com
    affiliation: Genentech
  - name: Di Cook
    email: di.cook@monash.edu
    affiliation: Monash University
address:
  - code: Monash University
    address: Department of Econometrics and Business Statistics, Clayton, Victoria, Australia
  - code: Genentech
    address: Bioinformatics and Computational Biology, Genentech, Inc., South San Francisco, California, United States of America 
abstract: |
    The Bioconductor project has created many useful data abstractions for analysing high-throughput genomics experiments. However, there is a cognitive load placed on a user in learning a data abstraction and understanding its appropriate use. Throughout a standard workflow, a user must navigate and know many of these abstractions to perform an genomic analysis task, when a single data abstraction, a GRanges object will suffice. The GRanges class naturally represent genomic intervals and their associated measurements. By recognising that the GRanges class follows 'tidy' data principles we have created a grammar of genomic data transformation. The grammar defines verbs for performing actions on and between genomic interval data. It provides a principled way
    of performing common genomic data analysis tasks through a coherent interface to existing Bioconductor infrastructure, resulting in human readable analysis workflows. We have implemented this grammar as an Bioconductor/R package called plyranges.

@@

By definition GRanges follows the tidy data pattern: it is a rectangular 
table corresponding to a single biological context. Each row contains a
single observation and each column is a variable about that
observation. Hence, we have designed the `plyranges` DSL to extend the grammar
and design principles of `dplyr`:  cohesion, consistency, endomorphism,
and fluency. All of these principles are defined and discussed
below in the context of the GRanges class. Where applicable we contrast our 
design to the existing Bioconductor infrastructure.

@@

The `plyranges` DSL defines an expressive algebra for performing genomic
arithemtic withn and between GRanges objects. Often in analyses of
genomic regions we are required to make adjustments to the start, end,
and width columns. This is supported directly in `plyranges` via extending
the `mutate()` function in `dplyr` (table 1). However, often adjustments
need to be relative: there is a mutual dependence between the start, end and
width columns over a genomic region. Changing the start,column needs to change
either the width or the end to preserve integrity of the object. The GRanges 
object by design, causes a user to think of the range as a pair of endpoints, 
and hence we have developed operators that perform coordinate transfomations with 
the notion of an 'anchoring' combinator.

@@

Binary operations between two ranges mostly align with set
operations that treat ranges as sets of integers. These are
parallel/vectorized operations between two vectors/tables of
ranges, and are not to be confused with the aggregating 
operations defined in table x. We make the binary nature of these operators
by having in-fix functions in R for the parallel union, intersect and
asymetric set difference. There are also operations that relate to the 
relative positions of two ranges such as the 'between' operator which
returns the range in the gap between two ranges and the 'span' operator 
a range that spans both ranges. The binary operations can be expressed within
the functions in table x.

@@

```{r olaps-fig, echo = FALSE, out.width="400pt", fig.cap="Illustration of the three overlap join operators. Each join takes query and subject range as input (black and grey rectangles, respectively) with associated metadata (pink and orange rectangles respectively). An index for the join is computed via returning a Hits object, which contains the indices of where the subject overlaps the query range. This index is used to expand the query or (take the intersect) of the query ranges with subject ranges and return metadata associated with each input. This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
knitr::include_graphics("olaps-figure.png")
```
We extend this idea by introducing three types of overlap joins: inner,
intersect and left (figure \ref{fig:olaps-fig}). The inner join considers a
hit to be any overlap between subject and query and expands the query range to 
have length equal to the  number of hits. The intersect join uses the same 
definition of a hit as the inner join but returns a range where the start and end coordinates
are the intersect of coordinates in the subject that hit the query. Finally,
the overlap left join is akin to left outer join in Cobb's relational algebra:
it performs an overlap inner join but also returns all query ranges that
are not hit by the subject.

@@

* compute coverage as an example here

This design pattern strongly deviates from the design of the OO interface in 
`GenomicRanges`, where many methods return a new class that is unfamiliar to 
the user upon return. These low-level classes enable efficient computing at 
the cost of complexity and are abstracted away in `plyranges` interface.

@@

```{r, echo = FALSE, eval = FALSE}
feature_tbl <- tribble(
  ~Verb, ~Action, ~Description,
  "`mutate()`", "modifies columns", "Takes a GRanges object and a set of name-value pairs and generates a new GRanges object that with modified or new metadata columns or modified core components (start, end, width, seqnames, strand).",
 "`filter()`", "subset rows", "Takes a GRanges object and a set of logical expressions and restricts the GRanges object to where all logical expressions are true.",
 "`summarise()`", "aggregate one or more columns by a function", "Takes a GRanges object and a set of name-value pairs and aggregates the GRanges according to functions evaluated in the name-value pairs.",
"`group_by()`", "partition a GRanges object by a column", "Create an implicit grouping of GRanges object by one or more columns. This modifies the actions of verbs, so they are performed on each partition created by the grouping.",
"`select()`", "select columns of a GRanges objects", "Drop or reorder metadata columns.",
"`arrange()`", "sort a GRanges object by columns", "Sort a GRanges object by named variables",
"`disjoin_ranges()`", "aggregate one more columns of a GRanges object over the union of end coordinates", "Takes a GRanges object and a set of name-value pairs and expands the GRanges object by taking the union of end coordinates. For each row that was included to form the final union, the aggregation function defined by a name-value pair is called.",
"`reduce_ranges()`", "aggregate one more columns of a GRanges object by merging overlapping and neighbouring ranges.", "Takes a GRanges object and a set of name-value pairs and reduces the GRanges object by merging overlapping and neighouring ranges. For each row that was included to form merged range, the aggregation function defined by a name-value pair is called."
)

knitr::kable(feature_tbl, format = "latex")
```

@@

First we filter out any insertion or deletion alleles and 
the SNPs present on the mitochondria then create a logical vector corresponding
to whether there is a transition event. 
```{r}
h1_snp_array <- h1_snp_array %>%
  filter(!(ref %in% c("I", "D")), seqnames != "M") %>%
  mutate(transition = (ref %in% c("A", "G") & alt %in% c("G","A")) |
                      (ref %in% c("C","T") & alt %in% c("T", "C")))
```

We can then compute the transition-transversion ratio using the `group_by`
and `summarise` pattern, it is computed as the total number of transition
SNPs divided by the total number of transversion SNPs within each chromsome
(figure \ref{fig:titv-viz}).

@@

@@

Note that the BAM file is only read into memory once we perform an operation
on it. Next we generate our genomic bins using the `tile` function
from `GenomicRanges`:

@@

We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall 
Institute and Dr Paul Harrison for their feedback on earlier drafts of
this work. We would also like to thank Lori Shepherd and Hèrve Pages
for the code review they performed. This report was written with
`knitr` @R-knitr and the figures were made with `ggbio` @R-ggbio. All code
required to reproduce this report is available at 
https://github.com/sa-lee/plyranges-paper.

>>>>

---
title: "plyranges: a grammar for transforming genomics data" 
author:
  - name: Stuart Lee
    email: stuart.lee1@monash.edu
    affiliation: Monash University
  - name: Michael Lawrence
    email: michael.lawrence@gene.com
    affiliation: Genentech
  - name: Di Cook
    email: di.cook@monash.edu
    affiliation: Monash University
address:
  - code: Monash University
    address: Department of Econometrics and Business Statistics, Clayton, Victoria, Australia
  - code: Genentech
    address: Bioinformatics and Computational Biology, Genentech, Inc., South San Francisco, California, United States of America 
abstract: |
    The Bioconductor project has created many useful data abstractions for analysing high-throughput genomics experiments. However, there is a cognitive load placed on a user in learning a data abstraction and understanding its appropriate use. Throughout a standard workflow, a user must navigate and know many of these abstractions to perform an genomic analysis task, when a single data abstraction, a GRanges object will suffice. The GRanges class naturally represent genomic intervals and their associated measurements. By recognising that the GRanges class follows 'tidy' data principles we have created a grammar of genomic data transformation. The grammar defines verbs for performing actions on and between genomic interval data. It provides a principled way
    of performing common genomic data analysis tasks through a coherent interface to existing Bioconductor infrastructure, resulting in human readable analysis workflows. We have implemented this grammar as a Bioconductor/R package called plyranges.

@@

By definition GRanges follows the tidy data pattern: it is a rectangular 
table corresponding to a single biological context. Each row contains a
single observation and each column is a variable about that
observation. Hence, we have designed the `plyranges` DSL to extend the grammar
and design principles of `dplyr`:  cohesion, consistency, endomorphism,
and fluency. All of these principles are defined and discussed
below in the context of the GRanges class and an overview of 
the grammar is provided in table \ref{tab:grammar}. 
Where applicable we contrast our design to the existing Bioconductor infrastructure.

\begin{table}[ht]
\label{tab:grammar}
\centering
\begin{tabular}{|l|l|p{4cm}|}
  \hline
Operation Type & Verb &  Description \\ 
  \hline
  Arithmetic &  `mutate()` & modifies any column \\
  Arithmetic & `select()` & select columns \\
  Arithmetic & `arrange()` & sort by columns \\
  Arithmetic & `set\_width()` & modifies width \\
  Arithmetic & `stretch()` & extend range by fixed amount \\
  Arithmetic & `shift\_(direction)` & shift coordinates \\
  Arithmetic & `flank\_(direction)` & generate flanking regions \\
  Arithmetic & `\%setop\%` & row-wise set operations \\
  Restriction & `filter()` & subset rows \\
  Restriction & `filter\_by\_overlaps()` & subset by overlap \\
  Restriction & `filter\_by\_non\_overlaps()` & subset by no overlap \\
  Merging & `join\_overlap\_()` & merge by overlapping ranges \\
  Merging & `join\_(follow/precedes/nearest)` & merge nearest neighbours \\
  Merging & `(setop)\_ranges` & range-wise set operations \\
  Aggregation & `summarise()` & aggregate over column(s) \\ 
  Aggregation & `disjoin\_ranges()` & aggregate column(s) over the union of end coordinates \\
  Aggregation & `reduce\_ranges()` & aggregate column(s) by merging overlapping and neighbouring ranges \\
  Modifier & `group\_by()` & partition by column(s)  \\ 
  Modifier & `group\_by\_overlaps()` & partition by overlaps \\
  Modifier & `anchor\_direction()` & fix coordinates \\
   \hline
\end{tabular}
\caption{Overview of `plyranges` grammar.}
\end{table}

@@

The `plyranges` DSL defines an expressive algebra for performing genomic
arithmetic with and between GRanges objects. Often in analyses of
genomic regions we are required to make adjustments to the start, end,
and width columns. This is supported directly in `plyranges` via extending
the `mutate()` function in `dplyr` (table \ref{tab:grammar}). 
However, often adjustments
need to be relative: there is a mutual dependence between the start, end and
width columns over a genomic region. Changing the start,column needs to change
either the width or the end to preserve integrity of the object. The GRanges 
object by design, causes a user to think of the range as a pair of endpoints, 
and hence we have developed operators that perform coordinate transformations with 
the notion of an 'anchoring' combinator.

@@

Binary operations between two ranges mostly align with set
operations that treat ranges as sets of integers. These are
parallel/vectorized operations between two vectors/tables of
ranges, and are not to be confused with the merging 
operations defined in table \ref{tab:grammar}. We make the binary nature of 
these operators
by having in-fix functions in R for the parallel union, intersect and
asymmetric set difference. There are also operations that relate to the 
relative positions of two ranges such as the 'between' operator which
returns the range in the gap between two ranges and the 'span' operator 
a range that spans both ranges. The binary operations can be expressed within
the arithmetic, restriction or aggregation functions in table \ref{tab:grammar}.

@@

```{r olaps-fig, echo = FALSE, out.width="400pt", fig.cap="Illustration of the three overlap join operators. Each join takes query and subject range as input (black and grey rectangles, respectively) with associated metadata (pink and orange rectangles respectively). An index for the join is computed via returning a Hits object, which contains the indices of where the subject overlaps the query range. This index is used to expand the query or (take the intersect) of the query ranges with subject ranges and return metadata associated with each input. This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
knitr::include_graphics("olaps-figure.png")
```

We extend this idea by introducing three types of overlap joins: inner,
intersect and left (figure \ref{fig:olaps-fig}). The inner join considers a
hit to be any overlap between subject and query and expands the query range to 
have length equal to the  number of hits. The intersect join uses the same 
definition of a hit as the inner join but returns a range where the start and end coordinates
are the intersect of coordinates in the subject that hit the query. Finally,
the overlap left join is akin to left outer join in Cobb's relational algebra:
it performs an overlap inner join but also returns all query ranges that
are not hit by the subject.

@@

As an example, in `plyranges` when we compute coverage over a GRanges object,
the result of the operation is an expanded GRanges object with a score column,
corresponding the estimated coverage over a genomic region. Hence, the resulting
coverage score can easily be composed with other expressions in our algebra.
This pattern strongly deviates from the design of the OO interface in 
`GenomicRanges`, which in the case of computing coverage would return an RleList,
an object that has many new methods and that is potentially unfamiliar to 
a new user. While, these low-level classes enable efficient computing,
the increase complexity, and as a consequence are abstracted away in the 
`plyranges` DSL.

@@

@@

First we filter out any insertion or deletion alleles and 
the SNPs present on the mitochondria then create a logical vector corresponding
to whether there is a transition event.

```{r}
h1_snp_array <- h1_snp_array %>%
  filter(!(ref %in% c("I", "D")), seqnames != "M") %>%
  mutate(transition = (ref %in% c("A", "G") & alt %in% c("G","A")) |
                      (ref %in% c("C","T") & alt %in% c("T", "C")))
```

We can then compute the transition-transversion ratio using the `group_by`
and `summarise` pattern, it is computed as the total number of transition
SNPs divided by the total number of transversion SNPs within each chromosome
(figure \ref{fig:titv-viz}).

@@

<!-- 
Trivial example? Could do something with a smooth summarised over
genomic coordinates? 
-->

@@

Note that the BAM file is only read into memory once we perform an operation
on it. Because of interoperability over Bioconductor, we can generate
genomic bins using the `tile` function from `GenomicRanges`:

@@

We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall 
Institute and Dr Paul Harrison for their feedback on earlier drafts of
this work. We would also like to thank Lori Shepherd and Hèrve Pages
for the code review they performed. This report was written with
`knitr` @R-knitr and the figures were made with `ggbio` @R-ggbio. All code
required to reproduce this article is available at 
https://github.com/sa-lee/plyranges-paper.

### 2018-04-24 paper.tex
complete draft (minus final example)

By definition GRanges follows the tidy data pattern: it is a rectangular
table corresponding to a single biological context. Each row contains a
single observation and each column is a variable about that observation.
Hence, we have designed the \texttt{plyranges} DSL to extend the grammar
and design principles of \texttt{dplyr}: cohesion, consistency,
endomorphism, and fluency. All of these principles are defined and
discussed below in the context of the GRanges class. Where applicable we
contrast our design to the existing Bioconductor infrastructure.

@@

The \texttt{plyranges} DSL defines an expressive algebra for performing
genomic arithemtic withn and between GRanges objects. Often in analyses
of genomic regions we are required to make adjustments to the start,
end, and width columns. This is supported directly in \texttt{plyranges}
via extending the \texttt{mutate()} function in \texttt{dplyr} (table
1). However, often adjustments need to be relative: there is a mutual
dependence between the start, end and width columns over a genomic
region. Changing the start,column needs to change either the width or
the end to preserve integrity of the object. The GRanges object by
design, causes a user to think of the range as a pair of endpoints, and
hence we have developed operators that perform coordinate transfomations
with the notion of an `anchoring' combinator.

@@

Binary operations between two ranges mostly align with set operations
that treat ranges as sets of integers. These are parallel/vectorized
operations between two vectors/tables of ranges, and are not to be
confused with the aggregating operations defined in table x. We make the
binary nature of these operators by having in-fix functions in R for the
parallel union, intersect and asymetric set difference. There are also
operations that relate to the relative positions of two ranges such as
the `between' operator which returns the range in the gap between two
ranges and the `span' operator a range that spans both ranges. The
binary operations can be expressed within the functions in table x.

@@

We extend this idea by introducing three types of overlap joins: inner,
intersect and left. The inner join considers a hit to be any overlap
between subject and query and expands the query range to have length
equal to the number of hits. The intersect join uses the same definition
of a hit as the inner join but returns a range where the start and end
coordinates are the intersect of coordinates in the subject that hit the
query. Finally, the overlap left join is akin to left outer join in
Cobb's relational algebra: it performs an overlap inner join but also
returns all query ranges that are not hit by the subject.

@@

\begin{itemize}
\tightlist
\item
  compute coverage as an example here
\end{itemize}

This design pattern strongly deviates from the design of the OO
interface in \texttt{GenomicRanges}, where many methods return a new
class that is unfamiliar to the user upon return. These low-level
classes enable efficient computing at the cost of complexity and are
abstracted away in \texttt{plyranges} interface.

@@

We can then compute the transition-transversion ratio using the
\texttt{group\_by} and \texttt{summarise} pattern, it is computed as the
total number of transition SNPs divided by the total number of
transversion SNPs within each chromsome (figure \ref{fig:titv-viz}).

@@

Note that the BAM file is only read into memory once we perform an
operation on it. Next we generate our genomic bins using the
\texttt{tile} function from \texttt{GenomicRanges}:

@@

We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall
Institute and Dr Paul Harrison for their feedback on earlier drafts of
this work. We would also like to thank Lori Shepherd and Hèrve Pages for
the code review they performed. This report was written with
\texttt{knitr} {[}17{]} and the figures were made with \texttt{ggbio}
{[}18{]}. All code required to reproduce this report is available at
https://github.com/sa-lee/plyranges-paper.

>>>>

By definition GRanges follows the tidy data pattern: it is a rectangular
table corresponding to a single biological context. Each row contains a
single observation and each column is a variable about that observation.
Hence, we have designed the \texttt{plyranges} DSL to extend the grammar
and design principles of \texttt{dplyr}: cohesion, consistency,
endomorphism, and fluency. All of these principles are defined and
discussed below in the context of the GRanges class and an overview of
the grammar is provided in table \ref{tab:grammar}. Where applicable we
contrast our design to the existing Bioconductor infrastructure.

\begin{table}[ht]
\label{tab:grammar}
\centering
\begin{tabular}{|l|l|p{4cm}|}
  \hline
Operation Type & Verb &  Description \\ 
  \hline
  Arithmetic &  `mutate()` & modifies any column \\
  Arithmetic & `select()` & select columns \\
  Arithmetic & `arrange()` & sort by columns \\
  Arithmetic & `set\_width()` & modifies width \\
  Arithmetic & `stretch()` & extend range by fixed amount \\
  Arithmetic & `shift\_(direction)` & shift coordinates \\
  Arithmetic & `flank\_(direction)` & generate flanking regions \\
  Arithmetic & `\%setop\%` & row-wise set operations \\
  Restriction & `filter()` & subset rows \\
  Restriction & `filter\_by\_overlaps()` & subset by overlap \\
  Restriction & `filter\_by\_non\_overlaps()` & subset by no overlap \\
  Merging & `join\_overlap\_()` & merge by overlapping ranges \\
  Merging & `join\_(follow/precedes/nearest)` & merge nearest neighbours \\
  Merging & `(setop)\_ranges` & range-wise set operations \\
  Aggregation & `summarise()` & aggregate over column(s) \\ 
  Aggregation & `disjoin\_ranges()` & aggregate column(s) over the union of end coordinates \\
  Aggregation & `reduce\_ranges()` & aggregate column(s) by merging overlapping and neighbouring ranges \\
  Modifier & `group\_by()` & partition by column(s)  \\ 
  Modifier & `group\_by\_overlaps()` & partition by overlaps \\
  Modifier & `anchor\_direction()` & fix coordinates \\
   \hline
\end{tabular}
\caption{Overview of `plyranges` grammar.}
\end{table}

@@

The \texttt{plyranges} DSL defines an expressive algebra for performing
genomic arithmetic with and between GRanges objects. Often in analyses
of genomic regions we are required to make adjustments to the start,
end, and width columns. This is supported directly in \texttt{plyranges}
via extending the \texttt{mutate()} function in \texttt{dplyr} (table
\ref{tab:grammar}). However, often adjustments need to be relative:
there is a mutual dependence between the start, end and width columns
over a genomic region. Changing the start,column needs to change either
the width or the end to preserve integrity of the object. The GRanges
object by design, causes a user to think of the range as a pair of
endpoints, and hence we have developed operators that perform coordinate
transformations with the notion of an `anchoring' combinator.

@@

Binary operations between two ranges mostly align with set operations
that treat ranges as sets of integers. These are parallel/vectorized
operations between two vectors/tables of ranges, and are not to be
confused with the merging operations defined in table \ref{tab:grammar}.
We make the binary nature of these operators by having in-fix functions
in R for the parallel union, intersect and asymmetric set difference.
There are also operations that relate to the relative positions of two
ranges such as the `between' operator which returns the range in the gap
between two ranges and the `span' operator a range that spans both
ranges. The binary operations can be expressed within the arithmetic,
restriction or aggregation functions in table \ref{tab:grammar}.

@@

We extend this idea by introducing three types of overlap joins: inner,
intersect and left (figure \ref{fig:olaps-fig}). The inner join
considers a hit to be any overlap between subject and query and expands
the query range to have length equal to the number of hits. The
intersect join uses the same definition of a hit as the inner join but
returns a range where the start and end coordinates are the intersect of
coordinates in the subject that hit the query. Finally, the overlap left
join is akin to left outer join in Cobb's relational algebra: it
performs an overlap inner join but also returns all query ranges that
are not hit by the subject.

@@

As an example, in \texttt{plyranges} when we compute coverage over a
GRanges object, the result of the operation is an expanded GRanges
object with a score column, corresponding the estimated coverage over a
genomic region. Hence, the resulting coverage score can easily be
composed with other expressions in our algebra. This pattern strongly
deviates from the design of the OO interface in \texttt{GenomicRanges},
which in the case of computing coverage would return an RleList, an
object that has many new methods and that is potentially unfamiliar to a
new user. While, these low-level classes enable efficient computing, the
increase complexity, and as a consequence are abstracted away in the
\texttt{plyranges} DSL.

@@

We can then compute the transition-transversion ratio using the
\texttt{group\_by} and \texttt{summarise} pattern, it is computed as the
total number of transition SNPs divided by the total number of
transversion SNPs within each chromosome (figure \ref{fig:titv-viz}).

@@

Note that the BAM file is only read into memory once we perform an
operation on it. Because of interoperability over Bioconductor, we can
generate genomic bins using the \texttt{tile} function from
\texttt{GenomicRanges}:

@@

We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall
Institute and Dr Paul Harrison for their feedback on earlier drafts of
this work. We would also like to thank Lori Shepherd and Hèrve Pages for
the code review they performed. This report was written with
\texttt{knitr} {[}17{]} and the figures were made with \texttt{ggbio}
{[}18{]}. All code required to reproduce this article is available at
https://github.com/sa-lee/plyranges-paper.
